FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-12-07 manuscript.Rmd
update 2024-12-07 07:27:16.199069

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' *empirical content* [REF],
making them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR Theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory,
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
then this formal theory is not FAIR.
FAIR Theory is thus consistent with, but does not require, formal theory.
There is more to say about this in light of the *Accessibility* principle,
where we discuss theories' "ability to be understood" [@guestWhatMakesGood2024] and the *problem of translation* [@quineReasonsIndeterminacyTranslation1970].

@@

```{r}
data.frame(
  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
  "X-interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
) |>
papaja::apa_table()
```

>>>>

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' *empirical content* [REF],
making them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR Theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory,
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
then this formal theory is not FAIR.
FAIR Theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).

@@

```{r tabmeehl}
data.frame(
  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
  "X-interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
) |>
papaja::apa_table()
```
